---
title: 'Ebola: Reproduction number and projection'
author: "Pierre Nouvellet & Anne Cori"
date: '2018-05-24'
---
  
  ```{r options, include = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(knitr)

opts_chunk$set(collapse = TRUE)


```


This file demonstrate how to use the ebolaproj package.

The package is used to estimate:

- the reproduction over a fixed time window
- project incidence forward from this time window



# Core Concepts 

From the previous lecture, we will further develop these concepts:
  
  - Compartmental models
- Flow diagrams and writing down a model 
- Natural history of Ebola
- Control strategies 
- Introduction to model fitting

## Required packages


```{r eval = FALSE}
install.packages("deSolve", dep=TRUE)
install.packages("gridExtra", dep = TRUE)
```

# About this document

## Contributors

- Pierre Nouvellet & Anne Cori: initial version


## Legal stuff

**License**: [CC-BY](https://creativecommons.org/licenses/by/3.0/)
**Copyright**: Pierre Nouvellet & Anne Cori, 2018


